### 2022-11-21 introduction/introduction.tex
Added detailed outline for paragraph on examples of proteins

Life is a physical phenomenon. Despite the complexity of living things, all these processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify these details into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) Since life spans a scale from a single cell to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and even use the other macromolecules as raw material. Proteins are therefore the molecular machines that carry out a cell's activities.

Macromolecule is a loose term applied to molecules with high relative molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.

To illustrate these concepts,

>>>>

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify their details into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, a punnettSince life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

FOOTNOTE: Macromolecule is a loose term applied to molecules with high molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.

Some examples will illustrate the central role of proteins more clearly.
These examples should be accessible to a wide audience, so nothing too abstract. Ideally should relate to human physiology and disease.
    1. Hemoglobin and sickle cell
        A. Introduce red blood cells transport oxygen
        B. Hemoglobin is protein that binds oxygen.
            i. Physically interacts with a molecule of oxygen
        C. Sickle cell is disease caused by mutation in hemoglobin protein.
            i. Cells produce hemoglobin with small mistake which causes them to clump into chains
            ii. This distorts the shape of red blood cells into a sickle
            iii. Red blood cells do not flow as easily through blood vessels causing pain and oxygen deprivation
    2. Photopsins and color blindness
        A. Photopsins are light sensitive proteins in cells in the retina at the back of the eye
        B. Three different versions are sensitive to red, green, and blue light
        C. Color blindness is the result of cells missing one of these proteins
    3. Lactase and lactose intolerance
        A. Lactose is a sugar found in milk that requires specific protein, lactase, to digest
        B. Many humans who can digest milk products in childhood, lose this ability in adulthood suffer symptoms such as bloating and diarrhea after consuming dairy products
        C. They stop producing lactase, so the undigested sugars are broken down by bacteria in the colon

Despite performing this diverse range of functions, all proteins are made up of a set of twenty simple building blocks called amino acids. Each amino acid in a protein is bonded to the next, forming a linear chain like beads on a string.
